\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-02 chapters/stbf_struct.tex
Comments Marc

	\todo{Rename to LDA}
	\todo{Work Kronecker sum in}
	\todo{Add results over time vs toeplitzlda}
	\todo{Introductory part on LCMV beamforming to intro}

@@

	\begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/covs.eps}
		\caption{Different estimators of the covariance and inverse covariance
			of 100 epochs of data from \textit{Subject 01} for channels
			\textit{Fz}, \textit{Cz}, \textit{Pz}, and \textit{Oz} and time samples between 0.1s and 0.6s.
			Regularized estimators of the inverse covariance exhibit less extreme values and have a sparser structure.
			(\textbf{a,b}) Empirical covariance and inverse covariance matrices.
			(\textbf{c,d}) Shrunk covariance and inverse covariance matrices with $\alpha=0.14$ as
			determined by the closed-form LOOCV method. (\textbf{e,f}) Kronecker-Toeplitz
			structured covariance and inverse covariance matrices.
			(\textbf{g,h}) Spatial Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.
			(\textbf{i,j}) Temporal Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.}
		\label{fig:kronlda-covs}
	\end{figure}
	\todo{visualize kronecker sum}
	\todo{visualize kronpca decomp pvl}
	\todo{method pvl}

@@

  \begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/fpi.eps}
		\caption{Average cross-validated \textsc{stbf-struct} Accuracy using
			one trial per block over all 21 subjects
			relative to the number of iterations used to estimate the Kronecker-Toeplitz structured shrunk
			covariance. Error bars represent the first and third quartiles. The
			accuracy does not improve when using more than one iteration.
			(\textbf{a}) Results for 1, 2, and 5 trials using only the first block in each
			training fold for training. (\textbf{b}) Results
			for 1, 2, and 5 trials using
			all nine training blocks in the training folds.
		}
		\label{fig:iterations}
	\end{figure}

@@

  \begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/accuracy.eps}
		\caption{Accuracy of the different classifiers for all 21 subjects relative to the
			number of blocks available for training. One block consists of 135
			epochs and corresponds to 27 seconds of stimulation. Accuracies
			are shown for the evaluation settings averaging over 1, 2, and
			3 trials of testing stimuli.
			\autoref{fig:accuracy_ap} contains results for all numbers of trials.
			While \textsc{stbf-emp} is unstable when little training data
			are available,
			regularization of the covariance matrix (\textsc{stbf-shrunk} and
			\textsc{stbf-struct}) drastically improves performance.}
		\label{fig:accuracy}
	\end{figure}
	\begin{table}[h]
		\centering
		\input{chapters/stbf_struct_tests_1.tex}
		\caption{$p$-values calculated by one-sided paired Wilcoxon rank-sum test with
			Holm correction using one testing trial for different classifiers
			and levels of data availability. $p$-values $<0.05$ are considered significant
			and marked in grey.}
		\label{tab:p_values_1}

		\centering
		\input{chapters/stbf_struct_tests_2.tex}
		\caption{$p$-values as in \autoref{tab:p_values_1} averaging over two testing trials.}
		\label{tab:p_values_2}

		\centering
		\input{chapters/stbf_struct_tests_3.tex}
		\caption{$p$-values as in \autoref{tab:p_values_1} averaging over five testing trials.}
		\label{tab:p_values_5}
	\end{table}
	\unskip

@@

	\begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/training_time.eps}
		\caption{Median fold training time for different classifiers at different spatial
			resolution levels evaluated over all training folds for all subjects.
			Error bars represent standard deviation. The training time of \textsc{stbf-shrunk} increases more
			steeply with resolution compared to \textsc{stbf-struct}.
			All \textsc{stbf} classifiers can be trained significantly faster than
			x\textsc{dawn+rg}.}
		\label{fig:training_time}
	\end{figure}

@@

	\todo{Remove mentions of interpretability}

@@

  \todo{Include part kronlda, check notes in journal}
\todo{
LCMV struct might be to restrictive
Toeplitzlda
MEG paper}
%\begin{equation}
%	\mat{C} = \mat{S}\otimes\mat{T}
%\end{equation}
%Tapering, Toeplitz and shrinkage
%\subsection{The Kronecker sum covariance model}
%\begin{equation}
%	\mat{C} = \sum_k^{CS}\mat{S}\otimes\mat{T}
%\end{equation}
\todo{pvl, kronpca}

\todo{check chapters/structure paper}
\todo{check structure Leuven.ai poster}
\todo{extension to the TFR domain}

>>>>

@@

	\begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/covs.eps}
    \caption[Estimated covariance and inverse covariance.]{Different estimators of the covariance and inverse covariance
			of 100 epochs of data from \textit{Subject 01} for channels
			\textit{Fz}, \textit{Cz}, \textit{Pz}, and \textit{Oz} and time samples between 0.1s and 0.6s.
			Regularized estimators of the inverse covariance exhibit less extreme values and have a sparser structure.
			(\textbf{a,b}) Empirical covariance and inverse covariance matrices.
			(\textbf{c,d}) Shrunk covariance and inverse covariance matrices with $\alpha=0.14$ as
			determined by the closed-form LOOCV method. (\textbf{e,f}) Kronecker-Toeplitz
			structured covariance and inverse covariance matrices.
			(\textbf{g,h}) Spatial Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.
			(\textbf{i,j}) Temporal Kronecker factor of the Kronecker-Toeplitz structured shrunk estimator and its inverse.}
		\label{fig:kronlda-covs}
	\end{figure}

@@

  \begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/fpi.eps}
    \caption[Average cross-validated \textsc{stbf-struct} accuracy]{%
      Average cross-validated \textsc{stbf-struct} accuracy using
			one trial per block over all 21 subjects
			relative to the number of iterations used to estimate the Kronecker-Toeplitz structured shrunk
			covariance. Error bars represent the first and third quartiles. The
			accuracy does not improve when using more than one iteration.
			(\textbf{a}) Results for 1, 2, and 5 trials using only the first block in each
			training fold for training. (\textbf{b}) Results
			for 1, 2, and 5 trials using
			all nine training blocks in the training folds.
		}
		\label{fig:iterations}
	\end{figure}

@@

  \begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/accuracy.eps}
    \caption[Clasifier accuracy in function of available training data.]{%
      Accuracy of the different classifiers for all 21 subjects relative to the
			number of blocks available for training. One block consists of 135
			epochs and corresponds to 27 seconds of stimulation. Accuracies
			are shown for the evaluation settings averaging over 1, 2, and
			3 trials of testing stimuli.
			\autoref{fig:accuracy_ap} contains results for all numbers of trials.
			While \textsc{stbf-emp} is unstable when little training data
			are available,
			regularization of the covariance matrix (\textsc{stbf-shrunk} and
			\textsc{stbf-struct}) drastically improves performance.}
		\label{fig:accuracy}
	\end{figure}
	\begin{table}[h]
		\centering
		\input{chapters/stbf_struct_tests_1.tex}
    \caption[Statistical significance of differences in classifier performance (1 repetition).]{%
      Statistical significance of differences in classifier performance.
      $p$-values calculated by one-sided paired Wilcoxon rank-sum test with
			Holm correction using one testing trial for different classifiers
			and levels of data availability. $p$-values $<0.05$ are considered significant
			and marked in grey.}
		\label{tab:p_values_1}

		\centering
		\input{chapters/stbf_struct_tests_2.tex}
		\caption[Statistical significance of differences in classifier performance (2 repetitions).]{%
      $p$-values as in \autoref{tab:p_values_1} averaging over two testing trials.}
		\label{tab:p_values_2}

		\centering
		\input{chapters/stbf_struct_tests_3.tex}
    \caption[Statistical significance of differences in classifier performance (5 repetitions).]{%
      $p$-values as in \autoref{tab:p_values_1} averaging over five testing trials.}
		\label{tab:p_values_5}
	\end{table}
	\unskip

@@

	\begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/training_time.eps}
    \caption[Classifier training time.]{Median fold training time for different classifiers at different spatial
			resolution levels evaluated over all training folds for all subjects.
			Error bars represent standard deviation. The training time of \textsc{stbf-shrunk} increases more
			steeply with resolution compared to \textsc{stbf-struct}.
			All \textsc{stbf} classifiers can be trained significantly faster than
			x\textsc{dawn+rg}.}
		\label{fig:training_time}
	\end{figure}

@@

@@

\todo{
in conclusion: LCMV struct might be to restrictive
Toeplitzlda MEG paper}

\todo{in conclusion: extension to the TFR domain}
%\begin{equation}
%	\mat{C} = \mat{S}\otimes\mat{T}
%\end{equation}
%Tapering, Toeplitz and shrinkage
%\subsection{The Kronecker sum covariance model}
%\begin{equation}
%	\mat{C} = \sum_k^{CS}\mat{S}\otimes\mat{T}
%\end{equation}

### 2024-10-02 chapters/wcble.tex
Started patients paper
Comments hakim
Comments Marc

\section{Introduction}
\label{sec:wcble/intro}

@@

\subsection{\Ac{erp} latency jitter}

@@

\begin{figure}[t]
  \centering
  \input{figures/wcble/smearing.tikz.tex}
  \caption{The smearing effect. When obtaining an \ac{erp} template through
  averaging, high jitter causes a decrease in peak amplitude and a deformation
  of the shape of the waveform.}
  \label{fig:wcble/smearing}
\end{figure}

@@

The W\ac{cble} algorithm is presented in Algorithm~\ref{alg:wcble}.
Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using equation~\ref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall \quad n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  Schematic representation of the Woody Classifier Based Latency
  Estimation training phase.
  (1) The first-stage spatiotemporal binary classifier is trained on a set of
  epochs.
  (2) It is	then applied to time-shifted copies of the epochs to obtain scores
  and class probabilities	over time.
  (3) The medians of these probability distributions are assumed the class
  latencies.
  (4) The epochs are aligned to their corresponding	class latencies by
  shifting in time such that all latencies fall at the same moment.
  (5) The spatiotemporal classifier is then retrained on the aligned epochs
  for a	next iteration.
  (6) After the iterative process halts, the scores and	latencies obtained
  from the last iteration are used to train the second-stage
  classifier.
}{%
  fig:diagram-train
}
The process halts after a fixed amount of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in Figure~\ref{fig:diagram-eval} to obtain a
vector of scores over time as in~Section~\ref{sec:cble} and the estimated latencies as
in~Section~\ref{sec:robust-latency}.
\fullpagefig{
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
}{%
  Schematic representation of the Woody Classifier Based Latency
  Estimation test phase. (1) The
	first-stage	spatiotemporal binary classifier obtained from the training phase is
	applied to time-shifted copies of the epochs to obtain scores and class probabilities
	over time. (2) The medians of these probability distributions are assumed
	as the new class latencies. (3) The scores and class latencies are input to the
  trained second-stage classifier, which predicts the label of the epochs.
}{%
  fig:diagram-eval
}
\todo{Pr not italic}
\todo{re-insert comments}

@@

	\caption{Classifier-Based Latency Estimation}
	\label{alg:cble}
\end{algorithm}

@@

	\smallskip \hrule \smallskip
	\textsc{Evaluate}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\For{$m=1\ldots M$}%\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{m,s}^\mathrm{test} \gets
      f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\mat{Y}^\mathrm{test},
  \mathbf{s}_\mathrm{target}^\mathrm{test},
  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
	\caption{Classifier-Based Latency Estimation with Woody Iterations}
	\label{alg:wcble}
\end{algorithm}

@@

%A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-sine}.
%\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
%\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}
}{%
  Simulated target and non-target evoked data. Targets contain a sine wave
  pulse jittered by $\sigma$. 32 EEG channels with pink
  spatiotemporal noise were simulated using a forward head model, channel C3 is
  plotted here.
}{fig:wcble/sim-sine}
\todo{Properly align plots and label axes}

@@

\begin{figure}[!ht]
  \input{figures/wcble/fig_latency.pgf} \\
  \input{figures/wcble/fig_accuracy.pgf}
  \caption{align results}
  \label{fig:wcble/results/latency}
\end{figure}
\todo{properly label axes}
\todo{discuss results}

>>>>

\section{Introduction}
\label{sec:wcble/intro}
\todohsm{do'nt repeat in introduction, rather make line clear}

@@

\subsection{\Acs{erp} latency jitter}

@@

\begin{figure}[t]
  \centering
  \input{figures/wcble/smearing.tikz.tex}
  \caption[The smearing effect.]{The smearing effect. When obtaining an \ac{erp} template through
  averaging, high jitter causes a decrease in peak amplitude and a deformation
  of the shape of the waveform.}
  \label{fig:wcble/smearing}
\end{figure}

@@

The W\ac{cble} algorithm is presented in Algorithm~\ref{alg:wcble}.
Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using equation~\ref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall \quad n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  Schematic representation of the \ac{wcble} training phase.
}{%
  Schematic representation of the \ac{wcble} training phase.
  (1) The first-stage spatiotemporal binary classifier is trained on a set of
  epochs.
  (2) It is	then applied to time-shifted copies of the epochs to obtain scores
  and class probabilities	over time.
  (3) The medians of these probability distributions are assumed the class
  latencies.
  (4) The epochs are aligned to their corresponding	class latencies by
  shifting in time such that all latencies fall at the same moment.
  (5) The spatiotemporal classifier is then retrained on the aligned epochs
  for a	next iteration.
  (6) After the iterative process halts, the scores and	latencies obtained
  from the last iteration are used to train the second-stage
  classifier.
}{%
  fig:diagram-train
}
The process halts after a fixed amount of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in Figure~\ref{fig:diagram-eval} to obtain a
vector of scores over time as in~Section~\ref{sec:cble} and the estimated latencies as
in~Section~\ref{sec:robust-latency}.
\fullpagefig{
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}

}{%
  Schematic representation of the \ac{wcble} test phase.
}{%
  Schematic representation of the Woody Classifier Based Latency
  Estimation test phase. (1) The
	first-stage	spatiotemporal binary classifier obtained from the training phase is
	applied to time-shifted copies of the epochs to obtain scores and class probabilities
	over time. (2) The medians of these probability distributions are assumed
	as the new class latencies. (3) The scores and class latencies are input to the
  trained second-stage classifier, which predicts the label of the epochs.
}{%
  fig:diagram-eval
}
\todo{Pr not italic}
\todo{re-insert comments}

@@

  \caption{\Ac{cble}}
	\label{alg:cble}
\end{algorithm}

@@

	\smallskip \hrule \smallskip
	\textsc{Evaluate}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\For{$m=1\ldots M$}%\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{m,s}^\mathrm{test} \gets
      f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
			    \mathrm{median}\left[Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\mat{Y}^\mathrm{test},
  \mathbf{s}_\mathrm{target}^\mathrm{test},
  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
  \caption{\ac{wcble}}
	\label{alg:wcble}
\end{algorithm}

@@

%A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-sine}.
%\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
%\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}
}{%
  Simulated target and non-target evoked data.
}{%
  Simulated target and non-target evoked data. Targets contain a sine wave
  pulse jittered by $\sigma$. 32 EEG channels with pink
  spatiotemporal noise were simulated using a forward head model, channel C3 is
  plotted here.
}{fig:wcble/sim-sine}
\todo{Properly align plots and label axes}
\todo{decrease tick size}

@@

\begin{figure}[!ht]
  \input{figures/wcble/fig_latency.pgf} \\
  \input{figures/wcble/fig_accuracy.pgf}
  \caption{caption}
  \label{fig:wcble/results/latency}
\end{figure}
\todo{properly label axes}
\todo{uniform fonts with rest of document}
\todo{correct width}
\todo{figure for data generation}
